Follow-up Samples

### 2017-03-07 submission/manuscript_outline_20161024.Rmd
modify abstract
additional background changes

**Background:** Colorectal cancer (CRC) is a worldwide health problem and research suggests a correlation between the fecal bacterial microbiome and CRC.  This study tested the hypothesis that changes in the bacterial community occur after treatment for adenoma or carcinoma.  Specifically, we tried to identify components within the community that were different before and after removal of lesion (adenoma or carcinoma).

@@

In this study we tested the hypothesis that there are detectable changes to the bacterial community between pre- (initial) and post- (follow up) treatment of lesion.  We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).  We utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up samples.  Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether initial and follow up samples were changing and whether it was towards a normal community.  We also investigated the two models for similar important OTUs to identify the crucial OTUs involved with both classifying lesion or normal and initial versus follow up samples.  This study will provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.

@@

***Carcinoma Associated Bacteria:***
Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].  So we next examined whether there were changes in previously well described carcinoma associated OTUs, such as *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  We first observed that only a small percentage of those with adenoma or carcinoma were positive or had a relative abundance above 0.1% for any of these respective OTUs [Figure 2].  Despite this, those with carcinoma had a decrease in relative abundance from initial to follow up for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 2A]. In contrast, there was no difference in relative abundance in any of these OTUs when considering only those with adenoma [Figure 2B].

>>>>

**Background:** Colorectal cancer (CRC) is a worldwide health problem and research suggests a correlation between the fecal bacterial microbiome and CRC.  This study tested the hypothesis that treatment for adenoma or carcinoma result in changes to the bacterial community.  Specifically, we tried to identify components within the community that were different before and after removal of lesion (adenoma or carcinoma).

@@

In this study we tested the hypothesis that there are detectable changes to the bacterial community between pre- (initial) and post- (follow up) treatment of lesion.  We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).  We then investigated several important genera from oral microbes that have been suggested to be important in CRC pathogenesis [@flynn_metabolic_2016].  We next utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up samples.  Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether initial and follow up samples were changing and whether it was towards a normal community.  We also investigated the two models for similar important OTUs to identify the crucial OTUs involved with both classifying lesion or normal and initial versus follow up samples.  This study will provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.

@@

***Carcinoma Associated Bacteria:***
So we next examined whether there were changes in previously well described carcinoma associated OTUs, such as *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  We first observed that only a small percentage of those with adenoma or carcinoma were positive or had a relative abundance above 0.1% for any of these respective OTUs [Figure 2].  Despite this, those with carcinoma had a decrease in relative abundance from initial to follow up for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 2A]. In contrast, there was no difference in relative abundance in any of these OTUs when considering only those with adenoma [Figure 2B].
